### 2024-09-27 paper/paper.qmd
Drafted introduction and data sections of paper

---
title: "Emergency Medical Response in Toronto: An Analysis of Ambulance Dispatches"
# subtitle: "My subtitle if needed"
author: 
  - Muhammad Abdullah Motasim
thanks: "Code and data are available at: https://github.com/abdullah-motasim/Analyzing-Ambulance-Response-Times"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
table-of-contents: true
toc-depth: 2
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| 
# Package loading
# install.packages("modeest")
library(tidyverse)
library(dplyr)
library(here)
library(tibble)
library(opendatatoronto)
library(ggplot2)
library(knitr)
library(modeest)

# Read data
raw_data <- read_csv(here("data/raw_data/raw_ambulance_response_data.csv"))
cleaned_data <- read_csv(here("data/analysis_data/cleaned_ambulance_response_data.csv"))
```

Quorto can only read data we load into it, so if we have data in global enviornment, quorto isnt able to use it

use (here::here("/path/to/folder)) to force quorto to look for folder outside of quorto working directory

use citation("library name") to get citation for a library

ggplot(cleaned_data, aes(x=year))+ geom_bar(position = "dodge")+ theme_minimal()

ggplot(cleaned_data, aes(x=year, fill=(units_arrived_at_scene)))+ geom_bar(position = "dodge")+ theme_minimal()+ \# theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Introduction

Efficient ambulances response times play a key role in saving saving peoples lives, especially in severe and sudden health complications such as cardiac arrests, trauma, and motor vehicle accidents. A medical study from 2011 found for cardiac arrest a 1 minute reduction in response time improved a persons odds of survival by 24% @citeBMJ. Furthermore, Toronto is facing issues retaining Emergency Medical Services (EMS) workers @citeCBCParamedicRetention leading to increased call wait times @citeCBC911WaitTimes and lower number of ambulances available to be dispatched @citeCBCAmbulanceResponseTimes.

This paper hopes to analyze the number and types of dispatch requests to EMS services within Toronto to identify the reason for the reduced amount of ambulances available and the increased call wait times. This is achieved utilizing a data set procured from OpenDataToronto @citeOpenDataTorontoWebsite containing information on the number of dispatches made per year, the severity of these reports and the number of units dispatched. We utilize this information to determine \<talk about how we found that the number of calls was increasing each year and the so was the severity. Analyze the severity data to see of higher severity means more dispatchers and if so say that the increase in the severity led to more calls and also more units needed per call, coupled with ems workers facing shortages and burnout led to poor ems services available to Canadians. Say thus is important as experts can now better tackle the underlying issues causing delayed ems response times, saving lives.\>

The remainder of this paper is structured as follows. @sec-data discusses the datatypes included in the raw data, the cleaning process for the data, and reason for selecting the data set we did. @sec-results analyzes the trends and correlations present between different variables utilizing tabular and graphical means. @sec-discussion the results of @sec-results going into detail on what these results can tell us about the cause for delayed ems response time. Lastly, @sec-nextsteps discusses limitations within the data and steps for improvement.

# Data {#sec-data}

## Raw Data

The data utilized for the analysis was sourced from the Open Data Toronto website and was read into the paper using the opendatatoronto library @citeOpenDataToronto, and all analysis was performed using R @citeR and the following packages: tidyverse @citeTidyverse, janitor @citeJanitor, lubridate @citeLubridate, truncnorm @citeTruncnorm, purrr @citePurrr, dplyr @citeDplyr, here @citeHere, ggplot2, @citeGgplot2, and knitr @citeKnitr. The chosen data set is called "Paramedic Services Incident Data" @citeOpenDataTorontoDataset and contains information about ambulances dispatches sent within Toronto from 2017-2022, the data is refreshed annually and was last updated October 5, 2023. The data is sourced and published by Toronto Paramedic Services meaning they decide the type of data collected and the method of collection. This set was chosen due to the large amount of data it contained, with over 1.7 million entries over the 5 years recorded meaning we could analyze trends over multiple years.

The raw downloaded data is shown in @tbl-raw-data-1 and @tbl-raw-data-2, as you can see it contains a total of 9 variables:

-   **ID** - Incident ID number associated with the Computer Aided Dispatch (CAD) record.
-   **Dispatch_Time** - Time the first Paramedic unit was assigned to the incident.
-   **Incident_Type** - Type of incident reported; possible values are:
    -   '-' (no value)
    -   Airport Standby
    -   Emergency Transfer
    -   Fire
    -   Medical
    -   Motor Vehicle Accident
-   **Priority_Number** - Priority level determined using the Medical Priority Dispatch System (MPDS), (see @sec-appendixA for details).
-   **Units_Arrived_At_Scene** - Number of Paramedic units that arrived on scene of incident
-   **Forward_Sortation_Area** - General location of incident based on Postal Code Forward Sortation Area

-   **Field Name** - Gives names of field such as ID, Dispatch_Time, Incident_Type, etc.
-   **Desctiption/Definition** - Describes what each field name means
-   **Comments/Examples** - Provides additional information on field name descriptions

```{r}
#| label: tbl-raw-data-1
#| tbl-cap: First 6 entires of the Paramedic Services Incident Dataset
#| echo: false
#| warning: false
#| message: false

head_raw_data <- head(raw_data)
# Note there is a type in Description within the original data set itself
knitr::kable(head_raw_data[,1:5], "simple",   col.names = c('ID', 'Dispatch_Time', 'Incident_Type', 'Priority_Number', 'Units_Arrived_At_Scene'))
```

```{r}
#| label: tbl-raw-data-2
#| tbl-cap: First 6 entires of the Paramedic Services Incident Dataset (cont.)
#| echo: false
#| warning: false
#| message: false
# Table is broken into 2 sections to show all data on a single page
knitr::kable(head_raw_data[,6:9], "simple",   col.names = c( 'Forward_Sortation_Area', 'Field Name', 'Desctiption/Definition', 'Comments/Examples'))
```

## Cleaned Data

During the creation of the raw data set the Field Names, Desctiption/Definition, and Comments/Examples columns were added as part of the process combining the data for all 5 years. However, we don't use these columns within our numerical analysis so they were removed within the data cleaning to reduce the data size. Furthermore, the dispatch_time column was formatted from type dbl to type POSIXct and two new columns were added containing the year and month of the report in order to allow for efficient plotting of dates. Lastly, the cleaned data was tested to ensure no missing values and all given values fell within their predetermined values. The cleaned data is shown in @tbl-cleaned-data-1 and @tbl-cleaned-data-2, @fig-year-vs-dispatches showcases the total number of dispatches per year.

```{r}
#| label: tbl-cleaned-data-1
#| tbl-cap: First 6 entires of the cleaned ambulance response data
#| echo: false
#| warning: false
#| message: false

head_cleaned_data <- head(cleaned_data)
# Note there is a type in Description within the original data set itself
knitr::kable(head_cleaned_data[,1:5], "simple",   col.names = c('id', 'dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene'))
```

```{r}
#| label: tbl-cleaned-data-2
#| tbl-cap: First 6 entires of the cleaned ambulance response data (cont.)
#| echo: false
#| warning: false
#| message: false

# Table is broken into 2 sections to show all data on a single page
knitr::kable(head_cleaned_data[,6:8], "simple",   col.names = c( 'forward_sortation_area', 'year', 'month'))
```

```{r}
#| label: fig-year-vs-dispatches
#| fig-cap: Year vs. Total Number of Dispatches
#| echo: false
#| warning: false
#| message: false

cleaned_data |>
ggplot(aes(x=year))+ 
  xlab("Year")+
  ylab("# of Dispatches")+
  geom_bar(position = "dodge")+ 
  scale_x_continuous(breaks = seq(min(cleaned_data$year), max(cleaned_data$year), by = 1)) + 
  theme_minimal()

```

## Summary of Data
We are only interested in the trend in dispatches over the years and @fig-year-vs-dispatches shows that there isn't much change in total number of dispatches over years meaning any calculated statistic on dispatch_time would not be of use. Thus,the only data types for which we can compute useful statistics are incident_type, priority_number, and units_arrived_at_scene. @tbl-cleaned-data-code shows the summary of these calculated statistics with mean for continuous data and mode for categorical data.

```{r}
#| label: tbl-cleaned-data-code
#| tbl-cap: Summary of Incident Data Statistics
#| echo: false
#| warning: false
#| message: false

# calculate statistics
mode_incident_type <- mlv(cleaned_data$incident_type, method = "mfv")
mode_priority_number <- mlv(cleaned_data$priority_number, method = "mfv")
mean_units_at_scene <- mean(cleaned_data$units_arrived_at_scene)

# Create a data frame to show values
summary_table <- data.frame(
  Field_Name = c("Units Arrived at Scene", "Mode Incident Type", "Mode Priority Number"),
  Value = c(mean_units_at_scene, mode_incident_type, mode_priority_number)
)

# Print the table
kable(summary_table)
```

# Results {#sec-results}

# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

# Weaknesses and next steps {#sec-nextsteps}

Data doesn't have arrival times

\newpage

\appendix

# Appendix {.unnumbered}

## Appendix A {#sec-appendixA}

Information on priority levels of Medical Priority Dispatch System (MPDS) ratings was sourced from @citePriorityDocument. Below is the severity of each level.

| Priority \# |                                Description |
|-------------|-------------------------------------------:|
| 5           |                       Alpha (Low Priority) |
| 4           |                       Bravo (Mid Priority) |
| 3           |        Charlie (Possibly Life Threatening) |
| 1           |                   Delta (Life Threatening) |
| 9           |       Echo (Full Arrest or Imminent Death) |
| 11          |         Alpha1 (Most Urgency within Alpha) |
| 12          |          Alpha2 (Mid Urgency within Alpha) |
| 13          |        Alpha3 (Least Urgency within Alpha) |
| 14          | Code 2 (Non-emergency transport requests.) |

\newpage

# References

### 2024-09-27 scripts/02-data_cleaning.R
Removed unnessecary data

@@ -15,10 +15,11 @@ library(dplyr)
 raw_data <- read_csv("data/raw_data/raw_ambulance_response_data.csv")
 
 cleaned_data <- suppressWarnings(
-  raw_data |>
+  raw_data[,2:5] |>
   janitor::clean_names() |>
-  # Remove unneeded columns which describe data types
-  select(-field_name, -desctiption_definition, -comments_examples) |>
+  # Remove columns which are not used in analysis
+  # select(-forward_sortation_area, -field_name, -desctiption_definition, -comments_examples) |>
+    
   na.omit() |>
   # Convert date and time variables
   separate(col = dispatch_time,
@@ -39,10 +40,10 @@ cleaned_data <- suppressWarnings(
     second = ifelse(is.na(second), "00", second)) |>
   # create a new date column
   mutate(dispatch_time = lubridate::ymd_hms(paste(year, month, day, hour, minute, second, sep = "-"))) |>
-  # Remove unneeded columns which describe data types
-  select(-hour, -minute, -second, -day) |>
+  # Remove unneeded columns which describe date
+  select(-hour, -minute, -second, -day, -month) |>
   #Reorder columns
-  select(id, dispatch_time, incident_type, priority_number, units_arrived_at_scene, forward_sortation_area, everything()))
+  select(dispatch_time, incident_type, priority_number, units_arrived_at_scene, year))
 
 #### Test data ####
 
@@ -50,16 +51,15 @@ cleaned_data <- suppressWarnings(
 paste("Negative Number of Dispatchers Test: ", cleaned_data$units_arrived_at_scene |> min() < 0)
 
 # test for dates within range specified
-dates <- c(2017, 2018, 2019, 2020, 2021, 2022)
-paste("Dispatch Time out-of-range Test: ", any(!cleaned_data$dispatch_time %in% priority_levels))
+years <- c(2017, 2018, 2019, 2020, 2021, 2022)
+paste("Dispatch Time out-of-range Test: ", any(!cleaned_data$year %in% years))
 
 # test for priority level meets requirements
 priority_levels <- c(1, 3, 4, 5, 9, 11, 12, 13, 14)
 paste("Priority Level Outside of Range Test: ", any(!cleaned_data$priority_number %in% priority_levels))
 
 # test for missing values
-any_NA_values <- all(is.na(cleaned_data$id))
-any_NA_values <- all(is.na(cleaned_data$date))
[…]

### 2024-09-27 paper/paper.qmd
Drafted full paper

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "The Toronto Paramedic Services respond to all ambulance dispatch requests across the city.
+In recent years, there have been growing concerns about increased call wait times and a shortage of available ambulances.
+This paper analyzes the number and severity of dispatch requests over the years.
+Our findings indicate that longer wait times are primarily due to a higher frequency of severe emergency calls following the COVID-19 pandemic, coupled with a shortage of first responders."
 format: pdf
@@
 ```
-
-Quorto can only read data we load into it, so if we have data in global enviornment, quorto isnt able to use it
-
-use (here::here("/path/to/folder)) to force quorto to look for folder outside of quorto working directory
-
-use citation("library name") to get citation for a library
-
-ggplot(cleaned_data, aes(x=year))+ geom_bar(position = "dodge")+ theme_minimal()
-
-ggplot(cleaned_data, aes(x=year, fill=(units_arrived_at_scene)))+ geom_bar(position = "dodge")+ theme_minimal()+ \# theme(axis.text.x = element_text(angle = 45, hjust = 1))
+\newpage
 
@@
 This is achieved utilizing a data set procured from OpenDataToronto @citeOpenDataTorontoWebsite containing information on the number of dispatches made per year, the severity of these reports and the number of units dispatched.
-We utilize this information to determine \<talk about how we found that the number of calls was increasing each year and the so was the severity.
-Analyze the severity data to see of higher severity means more dispatchers and if so say that the increase in the severity led to more calls and also more units needed per call, coupled with ems workers facing shortages and burnout led to poor ems services available to Canadians.
-Say thus is important as experts can now better tackle the underlying issues causing delayed ems response times, saving lives.\>
+We utilize this information to determine that COVID-19 reduced the number of calls received by Toronto  Paramedic Services, but also caused EMS workers to change their careers.
+Paired with an increase in the number of life-threatening and possibly life-threatening incidents meant the city was under equipped with experienced first responders to deal with the complex situations.
+Overall, all these factors lead to an increase in call wait times and decreased amount of available ambulances.
 
@@
 @sec-results analyzes the trends and correlations present between different variables utilizing tabular and graphical means.
-@sec-discussion the results of @sec-results going into detail on what these results can tell us about the cause for delayed ems response time.
+@sec-discussion discusses the results of @sec-results going into detail on what these results can tell us about the cause for delayed ems response time.
 Lastly, @sec-nextsteps discusses limitations within the data and steps for improvement.
@@
 
-The data utilized for the analysis was sourced from the Open Data Toronto website and was read into the paper using the opendatatoronto library @citeOpenDataToronto, and all analysis was performed using R @citeR and the following packages: tidyverse @citeTidyverse, janitor @citeJanitor, lubridate @citeLubridate, truncnorm @citeTruncnorm, purrr @citePurrr, dplyr @citeDplyr, here @citeHere, ggplot2, @citeGgplot2, and knitr @citeKnitr.
+The data utilized for the analysis was sourced from the Open Data Toronto website and was read into the paper using the opendatatoronto library @citeOpenDataToronto, and all analysis was performed using R @citeR and the following packages: tidyverse @citeTidyverse, janitor @citeJanitor, lubridate @citeLubridate, truncnorm @citeTruncnorm, purrr @citePurrr, dplyr @citeDplyr, here @citeHere, ggplot2, @citeGgplot2, knitr @citeKnitr, and modeest @citeModeest.
 The chosen data set is called "Paramedic Services Incident Data" @citeOpenDataTorontoDataset and contains information about ambulances dispatches sent within Toronto from 2017-2022, the data is refreshed annually and was last updated October 5, 2023.
 The data is sourced and published by Toronto Paramedic Services meaning they decide the type of data collected and the method of collection.
-This set was chosen due to the large amount of data it contained, with over 1.7 million entries over the 5 years recorded meaning we could analyze trends over multiple years.
+Online research shows they likely collect the data electronically as ambulance dispatches are made dispatchers manually record information such as the type of incident, priority level, dispatch time, etc.
+this data is likely combined into the data set and released.Note, this set was chosen due to the large amount of data it contained, with over 1.7 million entries over the 5 years recorded meaning we could analyze trends over multiple years.
 
@@
 -   **ID** - Incident ID number associated with the Computer Aided Dispatch (CAD) record.
+
 -   **Dispatch_Time** - Time the first Paramedic unit was assigned to the incident.
+
 -   **Incident_Type** - Type of incident reported; possible values are:
+
     -   '-' (no value)
@@
     -   Motor Vehicle Accident
--   **Priority_Number** - Priority level determined using the Medical Priority Dispatch System (MPDS), (see @sec-appendixA for details).
+
+-   **Priority_Number** - Priority level determined using the Medical Priority Dispatch System (MPDS), (see @sec-appendix for details).
+
 -   **Units_Arrived_At_Scene** - Number of Paramedic units that arrived on scene of incident
+
 -   **Forward_Sortation_Area** - General location of incident based on Postal Code Forward Sortation Area
@@
 -   **Field Name** - Gives names of field such as ID, Dispatch_Time, Incident_Type, etc.
+
 -   **Desctiption/Definition** - Describes what each field name means
+
 -   **Comments/Examples** - Provides additional information on field name descriptions
@@
 
+Since the data set is rather large the main goal when cleaning the data was to reduce the size.
 During the creation of the raw data set the Field Names, Desctiption/Definition, and Comments/Examples columns were added as part of the process combining the data for all 5 years.
 However, we don't use these columns within our numerical analysis so they were removed within the data cleaning to reduce the data size.
-Furthermore, the dispatch_time column was formatted from type dbl to type POSIXct and two new columns were added containing the year and month of the report in order to allow for efficient plotting of dates.
-Lastly, the cleaned data was tested to ensure no missing values and all given values fell within their predetermined values.
-The cleaned data is shown in @tbl-cleaned-data-1 and @tbl-cleaned-data-2, @fig-year-vs-dispatches showcases the total number of dispatches per year.
+Also, the ID and forward_sortation-area were removed as they didn't provide us with any useful information.
+The cleaned data is shown in @tbl-cleaned-data-1.
 
@@
 # Note there is a type in Description within the original data set itself
-knitr::kable(head_cleaned_data[,1:5], "simple",   col.names = c('id', 'dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene'))
+knitr::kable(head_cleaned_data[,1:5], "simple",   col.names = c('dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene', 'year'))
 ```
 
+## Summary of Data
+
+As we are interested in understanding the reason behind the reduced ambulance availability the only data types for which we can compute useful statistics are: number of dispatches per year, incident_type, priority_number, and units_arrived_at_scene.
+@tbl-cleaned-data-summary shows a brief summary of the data columns we are interested in, the summary of these calculated statistics with mean for continuous data and mode for categorical data.
+The total number of dispatches per year is visualized in @fig-year-vs-priority.
+on average the Toronto Paramedic Service's received 285 000 calls or 780 calls per day, with an average of 1.1 units sent per call that is approximately 860 units dispatched per day.
+
 ```{r}
-#| label: tbl-cleaned-data-2
-#| tbl-cap: First 6 entires of the cleaned ambulance response data (cont.)
+#| label: tbl-cleaned-data-summary
+#| tbl-cap: Summary of Incident Data Statistics
 #| echo: false
@@
 
-# Table is broken into 2 sections to show all data on a single page
-knitr::kable(head_cleaned_data[,6:8], "simple",   col.names = c( 'forward_sortation_area', 'year', 'month'))
+# calculate statistics
+mode_incident_type <- mlv(cleaned_data$incident_type, method = "mfv")
+mode_priority_number <- mlv(cleaned_data$priority_number, method = "mfv")
+mean_units_at_scene <- mean(cleaned_data$units_arrived_at_scene)
+
+total_dispatches_per_year <- cleaned_data %>%
+  group_by(year) %>%
+  summarise(total_dispatches = n())
+mean_dispatches_per_year <- mean(total_dispatches_per_year$total_dispatches)
+
+# Create a data frame to show values
+summary_table <- data.frame(
+  Field_Name = c("Mean Dispatches per Year", "Mean Units Arrived at Scene", "Mode Incident Type", "Mode Priority Number"),
+  Value = c(mean_dispatches_per_year, mean_units_at_scene, mode_incident_type, mode_priority_number)
+)
+
+# Print the table
+kable(summary_table)
 ```
@@
 
-## Summary of Data
-We are only interested in the trend in dispatches over the years and @fig-year-vs-dispatches shows that there isn't much change in total number of dispatches over years meaning any calculated statistic on dispatch_time would not be of use.
-Thus,the only data types for which we can compute useful statistics are incident_type, priority_number, and units_arrived_at_scene.
-@tbl-cleaned-data-code shows the summary of these calculated statistics with mean for continuous data and mode for categorical data.
+# Results {#sec-results}
+
+@fig-year-vs-dispatches appears to show there isn't much increase in the number of dispatches over the years, but we have to keep in mind the scale is in the hundreds of thousands, overall there is an increase of 8 000 calls per year with a dip in number of calls during the COVID-19 pandemic causing people to delay medical care @citeAvoidingCare.
+Overall, roughly 860 units were dispatched per day within the city and as of 2024 the Toronto Paramedic Services has a fleet of 236 ambulances @citeParamedicReport meaning each ambulance would need to be used 3.5 times per day to meet the demand explaining the lack of ambulances available for the city to use.
+@fig-year-vs-priority shows the change in priority numbers of the calls over the years, as seen priority numbers 1 and 3 increase and the rest decrease or stay the same near the end of 2022.
+Also, it is worth noting I have chosen to drop low priority numbers like: 11, 12, 13, and 14 as they occur less than 40 times which is a minuscule amount compared with the rest of the data.
 
 ```{r}
-#| label: tbl-cleaned-data-code
-#| tbl-cap: Summary of Incident Data Statistics
+#| label: fig-year-vs-priority
+#| fig-cap: Year vs. Priority Number
 #| echo: false
@@
 
-# calculate statistics
-mode_incident_type <- mlv(cleaned_data$incident_type, method = "mfv")
-mode_priority_number <- mlv(cleaned_data$priority_number, method = "mfv")
-mean_units_at_scene <- mean(cleaned_data$units_arrived_at_scene)
+# Calculate number of times each priority number appears each year
+priority_count_per_year <- cleaned_data %>%
+  filter(!priority_number %in% c(11,12,13,14)) %>%  # Exclude specific priority numbers
+  group_by(year, priority_number) %>%
+  summarise(count = n(), .groups = 'drop') %>%  # Summarize the count and drop extra grouping
+  arrange(year, priority_number)
 
-# Create a data frame to show values
-summary_table <- data.frame(
-  Field_Name = c("Units Arrived at Scene", "Mode Incident Type", "Mode Priority Number"),
-  Value = c(mean_units_at_scene, mode_incident_type, mode_priority_number)
-)
+# Plot years vs priority number
+ggplot(priority_count_per_year, aes(x = year, y = count, color = as.factor(priority_number), group = priority_number)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  labs(title = "Change in Priority Number Over the Years",
+       x = "Year",
+       y = "Count of Priority Number",
+       color = "Priority Number") +
+  theme_minimal() +
+  scale_x_continuous(breaks = unique(priority_count_per_year$year))
 
-# Print the table
-kable(summary_table)
 ```
 
-# Results {#sec-results}
+@fig-year-vs-1-unit-arrived and @fig-year-vs-2-units-arrived show how the number of units dispatched per year changed, with an overall positive trend until 2020 when the number dropped drastically inline with COVID-19 as mentioned previously.
+After which the initial drop numbers jump back up and resume the positive trend.
+It is worth noting we only look at 1 and 2 units dispatched as those are by far the most common dispatch numbers and we can see they both follow a similar trend.
+
+```{r}
+#| label: fig-year-vs-1-unit-arrived
+#| fig-cap: Year vs. Number of Times 1 Unit Arrived at Scene
+#| echo: false
+#| warning: false
+#| message: false
+
+# Calculate number of times each priority number appears each year
+units_count_per_year <- cleaned_data %>%
+  group_by(year, units_arrived_at_scene) %>%
+  summarise(count = n()) %>%
+  filter(count>100000) %>%
+  arrange(year, units_arrived_at_scene)
+
+# Plot years vs priority number
+ggplot(units_count_per_year, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  labs(title = "Change in Units Dispatched Over the Years",
+       x = "Year",
+       y = "Units Arrived at Scene",
+       color = "Units Arrived at Scene") +
+  theme_minimal() +
+  scale_x_continuous(breaks = unique(units_count_per_year$year))
+```
+
+```{r}
+#| label: fig-year-vs-2-units-arrived
+#| fig-cap: Year vs. Number of Times 2 Units Arrived at Scene
+#| echo: false
+#| warning: false
+#| message: false
+units_count_per_year2 <- cleaned_data %>%
+  group_by(year, units_arrived_at_scene) %>%
+  summarise(count = n()) %>%
+  filter(count<100000 & count>10000) %>%
+  arrange(year, units_arrived_at_scene)
+
+ggplot(units_count_per_year2, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  labs(title = "Change in Units Dispatched Over the Years",
+       x = "Year",
+       y = "Units Arrived at Scene",
+       color = "Units Arrived at Scene") +
+  theme_minimal() +
+  scale_x_continuous(breaks = unique(units_count_per_year$year))
+```
 
@@
 
-## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-## Third discussion point
+We have seen from the results section that there was a significant drop in the number of calls Toronto Paramedic Services received during the pandemic and a steady increase afterwards.
+However, this doesn't explain the increased wait time as the number of calls is still less than pre-COVID, but if we take into account the full context with many health care workers feelings burned and considering changing career paths @citeStatsCan.
+This coupled with the increase in calls and units dispatched after the COVID-19 pandemic ended resulted in the Toronto Paramedic Services being under equipped to deal with the increase in call volume.
+Furthermore, the increase in reports of priority 1 and 3, which are life-threatening or possibly life-threatening situations means that there was a need for experienced medical personnel at a time when there was a shortage.
+All these factors combined most likely resulted in longer wait times for Torontonians in need of medical assistance.
 
@@
 
-Data doesn't have arrival times
-
+A possible next step could be including the dispatch arrival time and seeing if there is any change within the time paramedic units take to respond to a request.
+As the data set I analyzed only contained the dispatch time I wasn't able to perform this calculation, however if completed we could get information on if the increased call wait times resulted in increased response times aswell.
 \newpage
@@
 
-# Appendix {.unnumbered}
-
-## Appendix A {#sec-appendixA}
-
+# Appendix {#sec-appendix}
 Information on priority levels of Medical Priority Dispatch System (MPDS) ratings was sourced from @citePriorityDocument. Below is the severity of each level.

### 2024-09-27 paper/paper.qmd
Updated draft

@@
 title: "Emergency Medical Response in Toronto: An Analysis of Ambulance Dispatches"
-# subtitle: "My subtitle if needed"
 author: 
@@
 
-Efficient ambulances response times play a key role in saving saving peoples lives, especially in severe and sudden health complications such as cardiac arrests, trauma, and motor vehicle accidents.
-A medical study from 2011 found for cardiac arrest a 1 minute reduction in response time improved a persons odds of survival by 24% @citeBMJ.
-Furthermore, Toronto is facing issues retaining Emergency Medical Services (EMS) workers @citeCBCParamedicRetention leading to increased call wait times @citeCBC911WaitTimes and lower number of ambulances available to be dispatched @citeCBCAmbulanceResponseTimes.
+Efficient ambulance response times play a key role in saving people's lives, especially in severe and sudden health complications such as cardiac arrests, trauma, and motor vehicle accidents.
+A medical study from 2011 found that for cardiac arrest a 1 minute reduction in response time improved a person's odds of survival by 24% @citeBMJ.
+Furthermore, Toronto is facing issues retaining Emergency Medical Services (EMS) workers @citeCBCParamedicRetention leading to increased call wait times @citeCBC911WaitTimes and a lower number of ambulances available to be dispatched @citeCBCAmbulanceResponseTimes.
 
 This paper hopes to analyze the number and types of dispatch requests to EMS services within Toronto to identify the reason for the reduced amount of ambulances available and the increased call wait times.
-This is achieved utilizing a data set procured from OpenDataToronto @citeOpenDataTorontoWebsite containing information on the number of dispatches made per year, the severity of these reports and the number of units dispatched.
+This is achieved utilizing a data set procured from OpenDataToronto @citeOpenDataTorontoWebsite containing information on the number of dispatches made per year, the severity of these reports, and the number of units dispatched.
 We utilize this information to determine that COVID-19 reduced the number of calls received by Toronto  Paramedic Services, but also caused EMS workers to change their careers.
-Paired with an increase in the number of life-threatening and possibly life-threatening incidents meant the city was under equipped with experienced first responders to deal with the complex situations.
-Overall, all these factors lead to an increase in call wait times and decreased amount of available ambulances.
+Paired with an increase in the number of life-threatening and possibly life-threatening incidents the city was under-equipped with experienced first responders to deal with the complex situations.
+Overall, all these factors lead to an increase in call wait times and a decreased amount of available ambulances.
 
 The remainder of this paper is structured as follows.
-@sec-data discusses the datatypes included in the raw data, the cleaning process for the data, and reason for selecting the data set we did.
-@sec-results analyzes the trends and correlations present between different variables utilizing tabular and graphical means.
+@sec-data discusses the data types included in the raw data, the cleaning process for the data, and the reason for selecting the data set we did.
+@sec-results analyzes the trends and correlations between different variables utilizing tabular and graphical means.
 @sec-discussion discusses the results of @sec-results going into detail on what these results can tell us about the cause for delayed ems response time.
@@
 
-The data utilized for the analysis was sourced from the Open Data Toronto website and was read into the paper using the opendatatoronto library @citeOpenDataToronto, and all analysis was performed using R @citeR and the following packages: tidyverse @citeTidyverse, janitor @citeJanitor, lubridate @citeLubridate, truncnorm @citeTruncnorm, purrr @citePurrr, dplyr @citeDplyr, here @citeHere, ggplot2, @citeGgplot2, knitr @citeKnitr, and modeest @citeModeest.
-The chosen data set is called "Paramedic Services Incident Data" @citeOpenDataTorontoDataset and contains information about ambulances dispatches sent within Toronto from 2017-2022, the data is refreshed annually and was last updated October 5, 2023.
-The data is sourced and published by Toronto Paramedic Services meaning they decide the type of data collected and the method of collection.
+The data utilized for the analysis was sourced from the Open Data Toronto website and was read into the paper using the opendatatoronto library @citeOpenDataToronto and all analysis was performed using R @citeR and the following packages: tidyverse @citeTidyverse, janitor @citeJanitor, lubridate @citeLubridate, truncnorm @citeTruncnorm, purrr @citePurrr, dplyr @citeDplyr, here @citeHere, ggplot2, @citeGgplot2, knitr @citeKnitr, and modeest @citeModeest.
+The chosen data set is called "Paramedic Services Incident Data" @citeOpenDataTorontoDataset and contains information about ambulance dispatches sent within Toronto from 2017-2022, the data is refreshed annually and was last updated on October 5, 2023.
+The data is sourced and published by Toronto Paramedic Services meaning they decide the type of data collected and the collection method.
 Online research shows they likely collect the data electronically as ambulance dispatches are made dispatchers manually record information such as the type of incident, priority level, dispatch time, etc.
-this data is likely combined into the data set and released.Note, this set was chosen due to the large amount of data it contained, with over 1.7 million entries over the 5 years recorded meaning we could analyze trends over multiple years.
+This data is likely combined into the data set and released.
+Note, that this set was chosen due to the large amount of data it contained, with over 1.7 million entries over the 5 years recorded meaning we could analyze trends over multiple years.
 
@@
 
--   **Dispatch_Time** - Time the first Paramedic unit was assigned to the incident.
+-   **Dispatch_Time** - The time the first Paramedic unit was assigned to the incident.
 
@@
 
--   **Units_Arrived_At_Scene** - Number of Paramedic units that arrived on scene of incident
+-   **Units_Arrived_At_Scene** - Number of Paramedic units that arrived on the scene of the incident
 
--   **Forward_Sortation_Area** - General location of incident based on Postal Code Forward Sortation Area
+-   **Forward_Sortation_Area** - General location of the incident based on Postal Code Forward Sortation Area
 
@@
 
--   **Desctiption/Definition** - Describes what each field name means
+-   **Description/Definition** - Describes what each field name means
 
@@
 head_raw_data <- head(raw_data)
-# Note there is a type in Description within the original data set itself
+# Note there is a typo in Description within the original data set itself
 knitr::kable(head_raw_data[,1:5], "simple",   col.names = c('ID', 'Dispatch_Time', 'Incident_Type', 'Priority_Number', 'Units_Arrived_At_Scene'))
@@
 #| message: false
+
 # Table is broken into 2 sections to show all data on a single page
@@
 
-Since the data set is rather large the main goal when cleaning the data was to reduce the size.
-During the creation of the raw data set the Field Names, Desctiption/Definition, and Comments/Examples columns were added as part of the process combining the data for all 5 years.
-However, we don't use these columns within our numerical analysis so they were removed within the data cleaning to reduce the data size.
-Also, the ID and forward_sortation-area were removed as they didn't provide us with any useful information.
-The cleaned data is shown in @tbl-cleaned-data-1.
+Since the data set is large, the main goal when cleaning the data was to reduce the size, meaning the cleaned data removed irrelevant columns from the raw data.
+The cleaned data shown in @tbl-cleaned-data-1  contains fewer columns than the raw data reducing the overall size significantly.
 
@@
 
-As we are interested in understanding the reason behind the reduced ambulance availability the only data types for which we can compute useful statistics are: number of dispatches per year, incident_type, priority_number, and units_arrived_at_scene.
-@tbl-cleaned-data-summary shows a brief summary of the data columns we are interested in, the summary of these calculated statistics with mean for continuous data and mode for categorical data.
+As we are interested in understanding the reason behind the reduced ambulance availability the only data types for which we can compute useful statistics are: the number of dispatches per year, incident_type, priority_number, and units_arrived_at_scene.
+@tbl-cleaned-data-summary shows a summary of the data columns we are interested in, the summary of these calculated statistics with mean for continuous data and mode for categorical data.
 The total number of dispatches per year is visualized in @fig-year-vs-priority.
-on average the Toronto Paramedic Service's received 285 000 calls or 780 calls per day, with an average of 1.1 units sent per call that is approximately 860 units dispatched per day.
+On average, the Toronto Paramedic Service received 285 000 calls or 780 calls per day, with an average of 1.1 units sent per call which is approximately 860 units dispatched daily.
 
@@
 
-@fig-year-vs-dispatches appears to show there isn't much increase in the number of dispatches over the years, but we have to keep in mind the scale is in the hundreds of thousands, overall there is an increase of 8 000 calls per year with a dip in number of calls during the COVID-19 pandemic causing people to delay medical care @citeAvoidingCare.
-Overall, roughly 860 units were dispatched per day within the city and as of 2024 the Toronto Paramedic Services has a fleet of 236 ambulances @citeParamedicReport meaning each ambulance would need to be used 3.5 times per day to meet the demand explaining the lack of ambulances available for the city to use.
+@fig-year-vs-dispatches appears to show there isn't much increase in the number of dispatches over the years.
+However, we have to keep in mind the scale is in the hundreds of thousands, overall there is an increase of 8 000 calls per year with a dip in the number of calls during the COVID-19 pandemic causing people to delay medical care @citeAvoidingCare.
+Overall, roughly 860 units were dispatched per day within the city and as of 2024, the Toronto Paramedic Services has a fleet of 236 ambulances @citeParamedicReport meaning each ambulance would need to be used 3.5 times per day to meet the demand explaining the lack of ambulances available for the city to use.
 @fig-year-vs-priority shows the change in priority numbers of the calls over the years, as seen priority numbers 1 and 3 increase and the rest decrease or stay the same near the end of 2022.
-Also, it is worth noting I have chosen to drop low priority numbers like: 11, 12, 13, and 14 as they occur less than 40 times which is a minuscule amount compared with the rest of the data.
+Also, it is worth noting I have chosen to drop low-priority numbers like 11, 12, 13, and 14 as they occur less than 40 times which is a minuscule amount compared with the rest of the data.
 
@@
 
-@fig-year-vs-1-unit-arrived and @fig-year-vs-2-units-arrived show how the number of units dispatched per year changed, with an overall positive trend until 2020 when the number dropped drastically inline with COVID-19 as mentioned previously.
-After which the initial drop numbers jump back up and resume the positive trend.
-It is worth noting we only look at 1 and 2 units dispatched as those are by far the most common dispatch numbers and we can see they both follow a similar trend.
+@fig-year-vs-1-unit-arrived and @fig-year-vs-2-units-arrived show how the number of units dispatched per year changed, with an overall positive trend until 2020 when the number dropped drastically in line with COVID-19 as mentioned previously.
+After which, the initial drop numbers jump back up and resume the positive trend.
+It is worth noting that we only look at 1 and 2 units dispatched as those are by far the most common dispatch numbers and we can see they both follow a similar trend.
 
@@
   summarise(count = n()) %>%
+  # Filter out low numbers
   filter(count>100000) %>%
@@
 #| message: false
+
+# Calculate number of times each priority number appears each year
 units_count_per_year2 <- cleaned_data %>%
@@
   summarise(count = n()) %>%
+  #Filter out low and extremely high numbers
   filter(count<100000 & count>10000) %>%
@@
 
+# Plot years vs priority number
 ggplot(units_count_per_year2, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
@@
 
-We have seen from the results section that there was a significant drop in the number of calls Toronto Paramedic Services received during the pandemic and a steady increase afterwards.
-However, this doesn't explain the increased wait time as the number of calls is still less than pre-COVID, but if we take into account the full context with many health care workers feelings burned and considering changing career paths @citeStatsCan.
-This coupled with the increase in calls and units dispatched after the COVID-19 pandemic ended resulted in the Toronto Paramedic Services being under equipped to deal with the increase in call volume.
+We have seen from the results section that there was a significant drop in the number of calls Toronto Paramedic Services received during the pandemic and a steady increase afterward. 
+However, this doesn't fully account for the increased wait times, as the volume of calls remains lower than pre-COVID levels.
+A more complete explanation considers the broader context, including many healthcare workers feeling burned out and contemplating career changes @citeStatsCan.
+This coupled with the increase in calls and units dispatched after the COVID-19 pandemic ended resulted in the Toronto Paramedic Services being under-equipped to deal with the increase in call volume.
 Furthermore, the increase in reports of priority 1 and 3, which are life-threatening or possibly life-threatening situations means that there was a need for experienced medical personnel at a time when there was a shortage.
-All these factors combined most likely resulted in longer wait times for Torontonians in need of medical assistance.
+All these factors combined most likely resulted in longer wait times for Torontonians requiring medical assistance.
 
@@
 A possible next step could be including the dispatch arrival time and seeing if there is any change within the time paramedic units take to respond to a request.
-As the data set I analyzed only contained the dispatch time I wasn't able to perform this calculation, however if completed we could get information on if the increased call wait times resulted in increased response times aswell.
+As the data set I analyzed only contained the dispatch time I couldn't perform this calculation, however, if completed we could get information on whether the increased call wait times resulted in increased response times as well.
+
 \newpage

### 2024-09-27 paper/paper.qmd
Updated draft

@@
 ---
-title: "Emergency Medical Response in Toronto: An Analysis of Ambulance Dispatches"
+title: "Increased Emergency Medical Response Wait Times: The Impact of Higher Call Volume and More Severe Cases Post-COVID-19"
 author: 

### 2025-06-18 scripts/00-simulate_data.R
converted download

@@ -1,69 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates a year of paramedic dispatch data
-# Author: Muhammad Abdullah Motasim
-# Date: 18 September 2024
-# Contact: abdullah.motasim@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: tidyverse, janitor, lubridate, truncnorm
-# Any other information needed: None
-
-
-#### Workspace setup ####
-set.seed(1009067563)
-library(tidyverse)
-library(janitor)
-library(lubridate)
-library(truncnorm)
-
-#### Simulate data ####
-
-# There are around 27 000  entries for each year
-# We will generate 30 000 entries to simulate data for the 2023 year
-size <- 30000
-
-# Create a sequence of dates and times within the 2023 year to choose from
-datetime_sequence <- seq(from = as.POSIXct("2023-01-01 00:00"), 
-                         to = as.POSIXct("2023-12-31 23:59"), 
-                         by = "min")
-
-# Format sequence to match data from Open Data Toronto
-datetime_sequence <- format(datetime_sequence, format = "%Y-%m-%d %H:%M")
-
-# Different calls have different priorities which are predetermined by the MDPS system
-# A weighted sampling is used to account for this disparity 
-priority_numbers = c(1, 3, 4, 5, 9, 11, 12, 13, 14)
-priority_numbers_weights <- c(0.15, 0.2, 0.3, 0.2, 0.1, 0.05, 0.04, 0.03, 0.03)
-
-# Different types of incidents occur more than others
-# A weighted sampling is used to account for this disparity 
-incident_types <- c("Medical", "Motor Vehicle Accident", "Emergency Transfer", 
-                    "Fire", "Airport Standby", "Other")
-incident_types_weights <- c(0.3, 0.3, 0.05, 0.1, 0.03, 0.2)
-
-# There are a different amount of dispatchers sent for each call
-# I assume there are typically 5 dispatchers sent per call
-# We sample from a truncated normal distribution to account for this disparity 
-max_dispactchers <- 22
-min_dispactchers <- 0
-mean <- 5
-sd <- 3
-  
-  
-simulated_data <-
-  tibble(
-    "id" = 1:size,
-    
-    "dispatch_time" = sample(x=datetime_sequence, size=size, replace=FALSE),
-
-    "incident_type" = sample(x = incident_types, size = size, replace = TRUE, 
-                             prob=incident_types_weights),
-    
[…]

### 2025-06-18 scripts/00-simulate_data.py

@@ -1,0 +1,49 @@
+#### Preamble ####
+# Purpose: Simulates a year of paramedic dispatch data (no scipy)
+# Author: Muhammad Abdullah Motasim (translated to Python)
+# Date: 18 September 2024
+
+import pandas as pd
+import numpy as np
+from datetime import datetime
+
+np.random.seed(1009067563)
+
+#### Simulate data ####
+
+size = 30000  # number of entries
+
+# Create a datetime sequence for every minute in 2023
+start = datetime(2023, 1, 1, 0, 0)
+end = datetime(2023, 12, 31, 23, 59)
+datetime_sequence = pd.date_range(start=start, end=end, freq="T").strftime("%Y-%m-%d %H:%M").tolist()
+
+# Priority levels and weights
+priority_numbers = [1, 3, 4, 5, 9, 11, 12, 13, 14]
+priority_weights = [0.15, 0.2, 0.3, 0.2, 0.1, 0.05, 0.04, 0.03, 0.03]
+
+# Incident types and weights
+incident_types = ["Medical", "Motor Vehicle Accident", "Emergency Transfer",
+                  "Fire", "Airport Standby", "Other"]
+incident_weights = [0.3, 0.3, 0.05, 0.1, 0.03, 0.2]
+
+# Simulate number of dispatchers using clipped normal distribution
+mean = 5
+sd = 3
+min_dispatchers = 0
+max_dispatchers = 22
+
+dispatchers = np.random.normal(loc=mean, scale=sd, size=size)
+dispatchers_clipped = np.clip(dispatchers, min_dispatchers, max_dispatchers).round().astype(int)
+
+# Build the dataset
+simulated_data = pd.DataFrame({
+    "id": np.arange(1, size + 1),
+    "dispatch_time": np.random.choice(datetime_sequence, size=size, replace=False),
+    "incident_type": np.random.choice(incident_types, size=size, p=incident_weights),
+    "priority_number": np.random.choice(priority_numbers, size=size, p=priority_weights),
+    "units_arrived_at_scene": dispatchers_clipped
+})
+
+# Save to CSV
+simulated_data.to_csv("data/raw_data/simulated_ambulance_response_data.csv", index=False)

### 2025-06-18 scripts/01-download_data.R

@@ -1,37 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from opendatatoronto
-# Author: Muhammad Abdullah Motasim
-# Date: 21 September 2024
-# Contact: abdullah.motasim@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: purrr, tidyverse, opendatatoronto
-# Any other information needed: None
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-library(purrr)
-
-# get package
-package <- show_package("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce")
-package
-
-#### Download data ####
-all_dispatch_data <-
-  list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce") |>
-  filter(name ==
-           "paramedic-services-incident-data-2017-2022") |>
-  get_resource()
-
-# Combine the data for all the years into a single dataframe
-combined_data <- bind_rows(all_dispatch_data)
-
-#### Save data ####
-
-write_csv(
-  x = combined_data,
-  file = "data/raw_data/raw_ambulance_response_data.csv"
-)
-
-head(combined_data)

### 2025-06-18 scripts/01-download_data.py

@@ -1,0 +1,42 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from opendatatoronto
+# Author: Muhammad Abdullah Motasim
+# Date: 21 September 2024
+# Contact: abdullah.motasim@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: purrr, tidyverse, opendatatoronto
+# Any other information needed: None
+
+import requests
+import pandas as pd
+from io import BytesIO
+
+print("Step 1: Requesting package metadata...")
+base_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca"
+url = base_url + "/api/3/action/package_show"
+params = { "id": "paramedic-services-incident-data" }
+package = requests.get(url, params=params).json()
+
+for resource in package["result"]["resources"]:
+    if not resource["datastore_active"] and resource["format"].lower() == "xlsx":
+        print(f"Step 2: Downloading resource: {resource['name']}")
+        resource_url = resource["url"]
+        response = requests.get(resource_url)
+
+        print("Step 3: Reading Excel file...")
+        xls = pd.ExcelFile(BytesIO(response.content))  # <-- likely bottleneck
+        print(f"Step 3.5: Found sheets: {xls.sheet_names}")
+
+        all_dfs = []
+        for sheet in xls.sheet_names:
+            print(f"Step 4: Reading sheet: {sheet}")
+            df = xls.parse(sheet)
+            print(f"Loaded {len(df)} rows from {sheet}")
+            all_dfs.append(df)
+
+        combined_df = pd.concat(all_dfs, ignore_index=True)
+        print("Step 5: Saving CSV...")
+        combined_df.to_csv("data/raw_data/raw_ambulance_data.csv", index=False)
+        print("Done.")
+        break
+            

### 2025-06-19 scripts/02-data_cleaning.R
finished changing scripts

@@ -1,69 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw ambulance dispatch response data
-# Author: Muhammad Abdullah Motasim
-# Date: 22 September 2024
-# Contact: abdullah.motasim@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: tidyverse, dplyr, raw_ambulance_response_data.csv
-# Any other information needed? None
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-
-#### Clean data ####
-raw_data <- read_csv("data/raw_data/raw_ambulance_response_data.csv")
-
-cleaned_data <- suppressWarnings(
-  raw_data[,2:5] |>
-  janitor::clean_names() |>
-  # Remove columns which are not used in analysis
-  # select(-forward_sortation_area, -field_name, -desctiption_definition, -comments_examples) |>
-    
-  na.omit() |>
-  # Convert date and time variables
-  separate(col = dispatch_time,
-           into = c("date", "time"),
-           sep = " ") |> 
-  # Separate date variable 
-  separate(col = date,
-           into = c("year", "month", "day"),
-           sep = "-") |>
-  # Separate time variable
-  separate(col = time,
-           into = c("hour", "minute", "second"),
-           sep = ":") |>
-  # Ensure we keep values which are at time 00:00:00
-  mutate(
-    hour = ifelse(is.na(hour), "00", hour),
-    minute = ifelse(is.na(minute), "00", minute),
-    second = ifelse(is.na(second), "00", second)) |>
-  # create a new date column
-  mutate(dispatch_time = lubridate::ymd_hms(paste(year, month, day, hour, minute, second, sep = "-"))) |>
-  # Remove unneeded columns which describe date
-  select(-hour, -minute, -second, -day, -month) |>
-  #Reorder columns
-  select(dispatch_time, incident_type, priority_number, units_arrived_at_scene, year))
-
-#### Test data ####
-
-# Test for no negative number of dispatchers
-paste("Negative Number of Dispatchers Test: ", cleaned_data$units_arrived_at_scene |> min() < 0)
-
-# test for dates within range specified
-years <- c(2017, 2018, 2019, 2020, 2021, 2022)
-paste("Dispatch Time out-of-range Test: ", any(!cleaned_data$year %in% years))
-
-# test for priority level meets requirements
[…]

### 2025-06-19 scripts/02-data_cleaning.py

@@ -1,0 +1,73 @@
+#### Preamble ####
+# Purpose: Cleans the raw ambulance dispatch response data
+# Author: Muhammad Abdullah Motasim
+# Date: 22 September 2024
+# Contact: abdullah.motasim@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: tidyverse, dplyr, raw_ambulance_response_data.csv
+# Any other information needed? None
+
+#### Workspace setup ####
+import pandas as pd
+import numpy as np
+
+#### Clean data ####
+raw_data = pd.read_csv("data/raw_data/raw_ambulance_response_data.csv")
+
+# Select columns 2 to 5 (Python uses 0-based indexing, so 1:5)
+data = raw_data.iloc[:, 1:5].copy()
+
+# Drop missing rows
+data = data.dropna()
+
+# Split 'dispatch_time' into date and time
+date_time_split = data["dispatch_time"].str.split(" ", expand=True)
+data["date"] = date_time_split[0]
+data["time"] = date_time_split[1]
+
+# Split date into year, month, day
+date_parts = data["date"].str.split("-", expand=True)
+data["year"] = date_parts[0]
+data["month"] = date_parts[1]
+data["day"] = date_parts[2]
+
+# Split time into hour, minute, second
+time_parts = data["time"].str.split(":", expand=True)
+data["hour"] = time_parts[0].fillna("00")
+data["minute"] = time_parts[1].fillna("00")
+data["second"] = time_parts[2].fillna("00")
+
+# Create dispatch_time column as datetime
+data["dispatch_time"] = pd.to_datetime(
+    data["year"] + "-" + data["month"] + "-" + data["day"] + " " +
+    data["hour"] + ":" + data["minute"] + ":" + data["second"],
+    errors="coerce"
+)
+
+# Keep and reorder necessary columns
+cleaned_data = data[[
+    "dispatch_time",
+    "incident_type",
+    "priority_number",
+    "units_arrived_at_scene",
+    "year"
+]]
+
+#### Test data ####
+
+# Test for no negative number of dispatchers
+print("Negative Number of Dispatchers Test: ", cleaned_data["units_arrived_at_scene"].min() < 0)
+
+# Test for dates within expected year range
+years = ['2017', '2018', '2019', '2020', '2021', '2022']
+print("Dispatch Time out-of-range Test: ", any(~cleaned_data["year"].isin(years)))
+
+# Test for priority level meets requirements
[…]

### 2025-06-19 scripts/03-test_data.R

@@ -1,34 +1,0 @@
-#### Preamble ####
-# Purpose: Sanity checks simulated response data
-# Author: Muhammad Abdullah Motasim
-# Date: 22 September 2024
-# Contact: abdullah.motasim@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: tidyverse, simulated_ambulance_response_data
-# Any other information needed: None
-
-
-#### Workspace setup ####
-library(tidyverse)
-data <- read_csv("data/raw_data/simulated_ambulance_response_data.csv")
-
-#### Test data ####
-
-# Test for no negative number of dispatchers
-paste("Negative Number of Dispatchers Test: ", data$units_arrived_at_scene |> min() < 0)
-
-# test for dates within range specified
-paste("Dispatch Time out-of-range Test: ", all(year(data$dispatch_time) != 2023))
-
-# test for priority level meets requirements
-priority_levels <- c(1, 3, 4, 5, 9, 11, 12, 13, 14)
-paste("Priority Level Outside of Range Test: ", any(!data$priority_number %in% priority_levels))
-
-# test for missing values
-any_NA_values <- all(is.na(data$id))
-any_NA_values <- all(is.na(data$dispatch_time))
-any_NA_values <- all(is.na(data$incident_type))
-any_NA_values <- all(is.na(data$priority_number))
-any_NA_values <- all(is.na(data$units_arrived_at_scene))
-
-paste("Any NA Values Test: ", any_NA_values)

### 2025-06-19 scripts/03-test_data.py

@@ -1,0 +1,41 @@
+#### Preamble ####
+# Purpose: Sanity checks simulated response data
+# Author: Muhammad Abdullah Motasim
+# Date: 22 September 2024
+# Contact: abdullah.motasim@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: tidyverse, simulated_ambulance_response_data
+# Any other information needed: None
+
+
+#### Workspace setup ####
+import pandas as pd
+import numpy as np
+
+data = pd.read_csv("data/raw_data/simulated_ambulance_response_data.csv")
+# Ensure dispatch_time is parsed as a datetime column
+data["dispatch_time"] = pd.to_datetime(data["dispatch_time"], errors="coerce")
+
+#### Test data ####
+
+# Test for no negative number of dispatchers
+print("Negative Number of Dispatchers Test: ", data["units_arrived_at_scene"].min() < 0)
+
+# Test for dates within range specified (no rows from calendar year 2023)
+print(
+    "Dispatch Time out-of-range Test: ",
+    (data["dispatch_time"].dt.year != 2023).all()
+)
+
+# Test for priority level meets requirements
+priority_levels = [1, 3, 4, 5, 9, 11, 12, 13, 14]
+print(
+    "Priority Level Outside of Range Test: ",
+    (~data["priority_number"].isin(priority_levels)).any()
+)
+
+# Test for missing values
+any_NA_values = data[
+    ["id", "dispatch_time", "incident_type", "priority_number", "units_arrived_at_scene"]
+].isna().any().any()
+print("Any NA Values Test: ", any_NA_values)

### 2025-06-19 paper/paper.qmd
translated paper code to python

@@
 
-```{r}
+```{python}
 #| include: false
@@
 #| message: false
-#| 
-# Package loading
-# install.packages("modeest")
-library(tidyverse)
-library(dplyr)
-library(here)
-library(tibble)
-library(opendatatoronto)
-library(ggplot2)
-library(knitr)
-library(modeest)
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
 
 # Read data
-raw_data <- read_csv(here("data/raw_data/raw_ambulance_response_data.csv"))
-cleaned_data <- read_csv(here("data/analysis_data/cleaned_ambulance_response_data.csv"))
+raw_data = pd.read_csv("../data/raw_data/raw_ambulance_response_data.csv")
+cleaned_data = pd.read_csv("../data/analysis_data/cleaned_ambulance_response_data.csv")
 ```
@@
 
-```{r}
+```{python}
 #| label: tbl-raw-data-1
@@
 
-head_raw_data <- head(raw_data)
-# Note there is a typo in Description within the original data set itself
-knitr::kable(head_raw_data[,1:5], "simple",   col.names = c('ID', 'Dispatch_Time', 'Incident_Type', 'Priority_Number', 'Units_Arrived_At_Scene'))
+head_raw_data = raw_data.head()
+
+# Select first 5 columns and rename them
+table = head_raw_data.iloc[:, 0:5].copy()
+table.columns = ['ID', 'Dispatch_Time', 'Incident_Type', 'Priority_Number', 'Units_Arrived_At_Scene']
+
+# Display the table (Quarto will render this as a formatted table)
+table
 ```
 
-```{r}
+```{python}
 #| label: tbl-raw-data-2
@@
 
-# Table is broken into 2 sections to show all data on a single page
-knitr::kable(head_raw_data[,6:9], "simple",   col.names = c( 'Forward_Sortation_Area', 'Field Name', 'Desctiption/Definition', 'Comments/Examples'))
+table_part2 = head_raw_data.iloc[:, 5:9].copy()
+table_part2.columns = ['Forward_Sortation_Area', 'Field Name', 'Desctiption/Definition', 'Comments/Examples']
+
+table_part2
 ```
@@
 
-```{r}
+```{python}
 #| label: tbl-cleaned-data-1
@@
 
-head_cleaned_data <- head(cleaned_data)
-# Note there is a type in Description within the original data set itself
-knitr::kable(head_cleaned_data[,1:5], "simple",   col.names = c('dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene', 'year'))
+head_cleaned_data = cleaned_data.head()
+
+# Select first 5 columns and rename
+table_cleaned = head_cleaned_data.iloc[:, 0:5].copy()
+table_cleaned.columns = ['dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene', 'year']
+
+table_cleaned
 ```
@@
 
-```{r}
+```{python}
 #| label: tbl-cleaned-data-summary
@@
 # calculate statistics
-mode_incident_type <- mlv(cleaned_data$incident_type, method = "mfv")
-mode_priority_number <- mlv(cleaned_data$priority_number, method = "mfv")
-mean_units_at_scene <- mean(cleaned_data$units_arrived_at_scene)
+# Mode for categorical column (incident_type)
+mode_incident_type = cleaned_data["incident_type"].mode().iloc[0]
 
-total_dispatches_per_year <- cleaned_data %>%
-  group_by(year) %>%
-  summarise(total_dispatches = n())
-mean_dispatches_per_year <- mean(total_dispatches_per_year$total_dispatches)
+# Mode for numerical column (priority_number)
+mode_priority_number = cleaned_data["priority_number"].mode().iloc[0]
+
+# Mean for numerical column (units_arrived_at_scene)
+mean_units_at_scene = cleaned_data["units_arrived_at_scene"].mean()
+
+# Total dispatches per year
+total_dispatches_per_year = (
+    cleaned_data.groupby("year")
+    .size()
+    .reset_index(name="total_dispatches")
+)
+mean_dispatches_per_year = total_dispatches_per_year["total_dispatches"].mean()
 
 # Create a data frame to show values
-summary_table <- data.frame(
-  Field_Name = c("Mean Dispatches per Year", "Mean Units Arrived at Scene", "Mode Incident Type", "Mode Priority Number"),
-  Value = c(mean_dispatches_per_year, mean_units_at_scene, mode_incident_type, mode_priority_number)
-)
+summary_table = pd.DataFrame({
+    "Field_Name": [
+        "Mean Dispatches per Year",
+        "Mean Units Arrived at Scene",
+        "Mode Incident Type",
+        "Mode Priority Number"
+    ],
+    "Value": [
+        round(mean_dispatches_per_year, 2),
+        round(mean_units_at_scene, 2),
+        mode_incident_type,
+        mode_priority_number
+    ]
+})
 
 # Print the table
-kable(summary_table)
+summary_table
 ```
 
-```{r}
+```{python}
 #| label: fig-year-vs-dispatches
@@
 
-cleaned_data |>
-ggplot(aes(x=year))+ 
-  xlab("Year")+
-  ylab("# of Dispatches")+
-  geom_bar(position = "dodge")+ 
-  scale_x_continuous(breaks = seq(min(cleaned_data$year), max(cleaned_data$year), by = 1)) + 
-  theme_minimal()
+# Count number of dispatches per year
+dispatch_counts = cleaned_data["year"].value_counts().sort_index()
+
+# Create numeric positions for bars
+x_pos = np.arange(len(dispatch_counts))  # e.g., [0, 1, 2, 3]
+
+# Plot
+plt.figure(figsize=(8, 5))
+plt.bar(x_pos, dispatch_counts.values)
+
+# Add axis labels and ticks
+plt.xlabel("Year")
+plt.ylabel("# of Dispatches")
+plt.xticks(ticks=x_pos, labels=dispatch_counts.index.astype(str), rotation=45, ha='right')
+
+# Add styling
+plt.title("")
+plt.grid(axis='y', linestyle='--', alpha=0.5)
+plt.tight_layout()
+plt.show()
 
@@
 
-```{r}
+```{python}
 #| label: fig-year-vs-priority
@@
 # Calculate number of times each priority number appears each year
-priority_count_per_year <- cleaned_data %>%
-  filter(!priority_number %in% c(11,12,13,14)) %>%  # Exclude specific priority numbers
-  group_by(year, priority_number) %>%
-  summarise(count = n(), .groups = 'drop') %>%  # Summarize the count and drop extra grouping
-  arrange(year, priority_number)
+priority_count_per_year = (
+    cleaned_data[~cleaned_data["priority_number"].isin([11, 12, 13, 14])]
+    .groupby(["year", "priority_number"])
+    .size()
+    .reset_index(name="count")
+    .sort_values(["year", "priority_number"])
+)
 
 # Plot years vs priority number
-ggplot(priority_count_per_year, aes(x = year, y = count, color = as.factor(priority_number), group = priority_number)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  labs(title = "Change in Priority Number Over the Years",
-       x = "Year",
-       y = "Count of Priority Number",
-       color = "Priority Number") +
-  theme_minimal() +
-  scale_x_continuous(breaks = unique(priority_count_per_year$year))
+plt.figure(figsize=(10, 6))
+
+for priority in priority_count_per_year["priority_number"].unique():
+    df_subset = priority_count_per_year[priority_count_per_year["priority_number"] == priority]
+    plt.plot(df_subset["year"], df_subset["count"], label=str(priority), marker='o', linewidth=1.5)
+
+plt.title("Change in Priority Number Over the Years")
+plt.xlabel("Year")
+plt.ylabel("Count of Priority Number")
+plt.xticks(sorted(priority_count_per_year["year"].unique()))
+plt.legend(title="Priority Number", title_fontsize=10)
+plt.grid(axis='y', linestyle='--', alpha=0.5)
+plt.tight_layout()
+plt.show()
 
@@
 
-```{r}
+```{python}
 #| label: fig-year-vs-1-unit-arrived
@@
 
-# Calculate number of times each priority number appears each year
-units_count_per_year <- cleaned_data %>%
-  group_by(year, units_arrived_at_scene) %>%
-  summarise(count = n()) %>%
-  # Filter out low numbers
-  filter(count>100000) %>%
-  arrange(year, units_arrived_at_scene)
+# Calculate number of times each units_arrived_at_scene value appears each year
+units_count_per_year = (
+    cleaned_data
+    .groupby(["year", "units_arrived_at_scene"])
+    .size()
+    .reset_index(name="count")
+)
 
-# Plot years vs priority number
-ggplot(units_count_per_year, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  labs(title = "Change in Units Dispatched Over the Years",
-       x = "Year",
-       y = "Units Arrived at Scene",
-       color = "Units Arrived at Scene") +
-  theme_minimal() +
-  scale_x_continuous(breaks = unique(units_count_per_year$year))
+# Filter out low-frequency combinations
+units_count_per_year = units_count_per_year[units_count_per_year["count"] > 100000]
+
+# Sort for consistency
+units_count_per_year = units_count_per_year.sort_values(["year", "units_arrived_at_scene"])
+
+# Plot years vs units_arrived_at_scene
+plt.figure(figsize=(10, 6))
+
+for units in units_count_per_year["units_arrived_at_scene"].unique():
+    df_subset = units_count_per_year[units_count_per_year["units_arrived_at_scene"] == units]
+    plt.plot(df_subset["year"], df_subset["count"], label=str(units), marker='o', linewidth=1.5)
+
+plt.title("Change in Units Dispatched Over the Years")
+plt.xlabel("Year")
+plt.ylabel("Units Arrived at Scene")
+plt.xticks(sorted(units_count_per_year["year"].unique()))
+plt.legend(title="Units Arrived at Scene", title_fontsize=10)
+plt.grid(axis='y', linestyle='--', alpha=0.5)
+plt.tight_layout()
+plt.show()
 ```
 
-```{r}
+```{python}
 #| label: fig-year-vs-2-units-arrived
@@
 
-# Calculate number of times each priority number appears each year
-units_count_per_year2 <- cleaned_data %>%
-  group_by(year, units_arrived_at_scene) %>%
-  summarise(count = n()) %>%
-  #Filter out low and extremely high numbers
-  filter(count<100000 & count>10000) %>%
-  arrange(year, units_arrived_at_scene)
+# Calculate number of times each units_arrived_at_scene value appears each year
+units_count_per_year2 = (
+    cleaned_data
+    .groupby(["year", "units_arrived_at_scene"])
+    .size()
+    .reset_index(name="count")
+)
 
-# Plot years vs priority number
-ggplot(units_count_per_year2, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  labs(title = "Change in Units Dispatched Over the Years",
-       x = "Year",
-       y = "Units Arrived at Scene",
-       color = "Units Arrived at Scene") +
-  theme_minimal() +
-  scale_x_continuous(breaks = unique(units_count_per_year$year))
+# Filter out low and extremely high values
+units_count_per_year2 = units_count_per_year2[
+    (units_count_per_year2["count"] < 100000) & (units_count_per_year2["count"] > 10000)
+]
+
+# Sort for consistent plotting
+units_count_per_year2 = units_count_per_year2.sort_values(["year", "units_arrived_at_scene"])
+
+# Plot years vs units_arrived_at_scene
+plt.figure(figsize=(10, 6))
+
+for units in units_count_per_year2["units_arrived_at_scene"].unique():
+    df_subset = units_count_per_year2[units_count_per_year2["units_arrived_at_scene"] == units]
+    plt.plot(df_subset["year"], df_subset["count"], label=str(units), marker='o', linewidth=1.5)
+
+plt.title("Change in Units Dispatched Over the Years")
+plt.xlabel("Year")
+plt.ylabel("Units Arrived at Scene")
+plt.xticks(sorted(units_count_per_year2["year"].unique()))
+plt.legend(title="Units Arrived at Scene", title_fontsize=10)
+plt.grid(axis='y', linestyle='--', alpha=0.5)
+plt.tight_layout()
+plt.show()
 ```
